---
title: "Solutions to lab 3 exercise 6"
author: "Bob Douma"
date: "13 November 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exercise 1.1**: Calculate the logarithm of the sequence 1:10, using first a for loop and then without a for loop.
```{r}
for (i in 1:10){
  a = log(i)
  print(a) # to show what you have done
}

log(1:10)

```


**Exercise 2.1**: Given some data created from the following code `c(25,1,10,89, NA, NA)`, 
calculate the mean value and the standard error of this mean ($s.e.m. = \sigma/\sqrt{n}$, where $\sigma$ is the standard deviation and $n$ is the amount of data) by ignoring missing values.

```{r}
a = c(25,1,10,89, NA, NA)
mean(a,na.rm=T)
sd(a,na.rm=T)/length(na.omit(a))
```

**Exercise 3.1**: Build a function to calculate the standard deviation ($\sigma = \sqrt{\frac{\sum_{i = 1}^n\left(x_i - \bar x\right)^2}{n - 1}}$). Test your function with some data that includes missing values.

```{r}
sigma.self = function(x,na.rm=F){
  mean = mean(x,na.rm=na.rm)
  n = length(na.omit(x))
  sd = sqrt(sum((x-mean)^2,na.rm=na.rm)/(n-1))
  return(sd)
}

```

**Exercise 4.1** *: Construct a curve
that has a maximum at ($x=5$, $y=1$).  Write the
equation, draw the curve in `R`, and explain
how you got there.

```{r}
# for example
curve(-(x-5)^2+1,from=0,to=10)
``` 

A quadratic functionh $x^2$ has a maximum. or minimum. This maximum/minimum is at $x=0$. To shift the curve to have its maximum at $x=5$, one need to subtract $5$ from the $x$, i.e. $(x-5)^2$. The maximum of this function is at $y=0$, so one need to add $1$ to the function, i.e. $(x-5)^2+1$.


## Below the solutions to Exercise 6 of Lab 3

1. The first step is to read in the data. The data is stored in different sheet of the file `shapes.xlsx`. Select a sheet, go to `file` and 'save as' and save the sheet as a comma-separated-file (.csv). Beware that if you are using a computer with dutch settings, you may run into problems for two reasons: First, the decimal character is a comma `,` in Dutch while a dot `.` in english. Second the defaul list separator is a semicolon `;` in Dutch while a `,` in English. Check the file that you saved to check the settings of your computer. If this is not correct; go to control panel; to region and language settings; to advanced settings and change the decimal character into a `.`  and the list separator (lijstschijdingsteken) in to a `,`.



```{r, eval=F}
setwd("D://...//...//...) # fill in location
shapes1 = read.csv(shapes1.csv)
```

2. After you have read in the data, you can make a plot through `plot(shapes1$y~ shapes1$x)`. Multiple plots can be made through specifying `par(mfrow=c(3,2))`. This will setup the grid, after using `plot` six times, the grid will be filled with plots.

3. Choosing appropriate deterministic functions

      **dataset 1**
      light response curve. There are a number of options of functions to choose from, depending on the level of sophistication:
      $\frac{ax}{(b+x)}$, $a(1-e^{(-bx)})$, $\frac{1}{2\theta}(\alpha I+p_{max}-\sqrt(\alpha I+p_{max})^2-4\theta I p_{max})$ see page 98 of Bolker.
      
      **dataset 2**
      The dataset describes a functional response. Bolker mentions four of those $min(ax,s)$ $\frac{ax}{(b+x)}$, $\frac{ax^2}{(b^2+x^2)}$,$\frac{ax^2}{(b+cx+x^2)}$
      
      **dataset 3**
      Allometric relationships have the form $ax^b$
      
      **dataset 4**
      This could be logistic growth $n(t)=\frac{K}{1+(\frac{K}{n_0})e^{-rt}}$ or the gompertz function $f(x)=e^{-ae^{-bx}}$
      
      **dataset 5**
      What about a negative exponential? $ae^{-bx}$ or a power function $ax^b$
      
      **dataset 6**
      Species reponse curves are curves that describe the probability of presence as a function of some factor. A good candidate good be a unimodel response curve. You could take the equation of the normal distribution without the scaling constant: e.g.
      $a e^{\frac{-(x-\mu)^2}{2\sigma^2}}$

4. See the word file on blackboard "Bestiary of functions.docx"

5. Curves can be added to the plots through `curve`: e.g. `curve(2x+2x,from=0,to=20)` 

      **dataset 1** 
      Reasonable values for the first dataset assuming a michaelis menten relationship are a=25 and b=60. For the            non-rectangular parabola one could choose values of theta = 0.7; a = 0.25; pmax = 25. 
      

      **dataset 2**
      `curve(ifelse(x>27,18,(2/3)*x),add=T)`
      
      `curve(20*x/(10+x),add=T)`
      
      **dataset 3**
      `curve(0.6*x^2,add=T)`

      **dataset 4**
        `K = 200; r = 0.2; N0=2; curve(K/(1+(K/N0)*exp(-r*x)),add=T)`
        
      **dataset 5**  
      `curve(8*(exp(-0.75*x)),add=T)  `
    
      **dataset 6**
      `mu = 5; b = 2; curve(exp(-(mu-x)^2/b)  ,add=T)`

      

