---
title: "Solutions to lab 3 exercise 6"
author: "Bob Douma"
date: "13 November 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Below the solutions to Exercise 6 of Lab 3

1. The first step is to read in the data. The data is stored in different sheet of the file `shapes.xlsx`. Select a sheet, go to `file` and 'save as' and save the sheet as a comma-separated-file (.csv). Beware that if you are using a computer with dutch settings, you may run into problems for two reasons: First, the decimal character is a comma `,` in Dutch while a dot `.` in english. Second the defaul list separator is a semicolon `;` in Dutch while a `,` in English. Check the file that you saved to check the settings of your computer. If this is not correct; go to control panel; to region and language settings; to advanced settings and change the decimal character into a `.`  and the list separator (lijstschijdingsteken) in to a `,`.



```{r, eval=F}
setwd("D://...//...//...) # fill in location
shapes1 = read.csv(shapes1.csv)
```

2. After you have read in the data, you can make a plot through `plot(shapes1$y~ shapes1$x)`. Multiple plots can be made through specifying `par(mfrow=c(3,2))`. This will setup the grid, after using `plot` six times, the grid will be filled with plots.

3. Choosing appropriate deterministic functions

      **dataset 1**
      light response curve. There are a number of options of functions to choose from, depending on the level of sophistication:
      $\frac{ax}{(b+x)}$, $a(1-e^{(-bx)})$, $\frac{1}{2\theta}(\alpha I+p_{max}-\sqrt(\alpha I+p_{max})^2-4\theta I p_{max})$ see page 98.
      
      **dataset 2**
      The dataset describes a functional responses. Bolker mentions four of those $min(ax,s)$ $\frac{ax}{(b+x)}$, $\frac{ax^2}{(b^2+x^2)}$,$\frac{ax^2}{(b+cx+x^2)}$
      
      **dataset 3**
      Allometric relationship have the form $ax^b$
      
      **dataset 4**
      This could be logistic growth $n(t)=\frac{K}{1+(\frac{K}{n_0})e^{-rt}}$ or the gompertz function $f(x)=e^{-ae^{-bx}}$
      
      **dataset 5**
      What about a negative exponential? $ae{-bx}$ or a power function $ax^b$
      
      **dataset 6**
      Species reponse curves are curves that describe the probability of presence as a function of some factor. A good candidate good be a unimodel response curve. You could take the equation of the normal distribution without the scaling constant: e.g.
      $a e^{\frac{-(x-\mu)^2}{2\sigma^2}}$

4. For example: `curve(2x+2x,from=0,to=20)` Reasonable values will be supplied in later tutorials